A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-01-17 ManuscriptGeophysics2018.tex
Typos

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
%
% (1) PRINCIPAL OBJECTIVES AND SCOPE OF THE WORK
  The open-source code \texttt{fdesign} makes it possible to design digital
  linear filters for the Hankel and Fourier transforms used in potential-,
  diffusive-, and wavefield modeling. Digital filters can be derived for any
  electromagnetic method, such as methods in the diffusive limits (DC, CSEM) as
  well as methods using higher frequency content (GPR, acoustic and elastic
  wavefields).
%
% (2) METHODOLOGY
  The direct matrix inversion method is used for the derivation of the filter
  values, and a brute-force inversion is carried out over the defined spacing
  and shifting values of the filter basis. Included or user-provided
  theoretical transform pairs are used for the inversion. Alternatively, one
  can provide 1D subsurface models that will be computed with a precise
  quadrature method using the EM modeler \texttt{empymod} to generate
  numerical transform pairs.
%
% (3) RESULTS
  The comparison of the presented 201\,pt filter with previously presented
  filters shows that it performs better for some standard CSEM cases. The
  derivation of a longer 2001\,pt filter for a GPR example with 250\,MHz centre
  frequency proves that the filter method works also for wave phenomena, not
  only for diffusive EM fields.
%
% (4) CONCLUSIONS
  The presented algorithm provides a tool to create problem specific digital
  filters. Such purpose-built filters can be made shorter and can speed up
  consecutive  potential-, diffusive-, and wavefield inversions.
%
\end{abstract}

@@

\cite{GP.71.Ghosh} states that the DLF idea is based on suggestions made four
decades earlier by \cite{PHY.33.Slichter} and \cite{GEO.40.Pekeris}, in that
``\emph{the kernel function is dependent only on the layer parameters, and an
expression relating it to the field measurements can be obtained by
mathematical processes.}'' However, until the introduction of DLF, these
suggestions found no widespread use, likely because of the missing computer
power to calculate the filter coefficients. He further states that credit goes
to his Ph.D. supervisor Otto \cite{BK.68.Koefoed, GP.70.Koefoed}, who retook
the task of direct interpretational methods with the introduction of raised
kernel functions. DLF is, as such, an improvement of that approach, providing a
faster and simpler method.

@@

The most well-known codes are likely Anderson's freely available ones.
\cite{USGS.73.Anderson} extends the method to transient responses, applying DLF
not only to the Hankel transform, but also to the Fourier transform. A
transient signal can therefore be obtained be applying twice a digital filter
to the wavenumber-frequency domain calculation. \cite{USGS.75.Anderson,
GEO.79.Anderson} presents improved filters for both Fourier and Hankel
transforms, introducing measures to significantly speed-up the calculation,
such as the lagged convolution or using the same abscissae for $J_0$ and $J_1$.
\cite{TMS.82.Anderson} and the included 801\,pt filter became the de facto
industry standard, to which subsequent filters were compared.
\cite{GEO.89.Anderson} presents a hybrid solution that uses either the DLF or a
much slower but highly accurate adaptive-quadrature approach, which can also be
used to measure the accuracy of the DLF. \cite{GEO.12.Key} presented codes for
comparing the DLF approach with a speed optimized quadrature method call
quadrature-with-extrapolation (QWE); despite its optimizations, QWE is still
not as fast as DLF, but it remains valuable as an independent tool for testing
the accuracy of particular filters for the DLF method. Other examples include
the codes by \cite{GP.75.Johansen}, an interactive system for interpretation of
resistivity soundings, and a tool to calculate filter coefficients by
\cite{GP.90.Christensen}. The latter is available upon request and was used,
for instance, in all the open-source modeling and inversion routines of CSIRO
in the Amira Project 223 \citep{ASEG.07.Raiche}.

@@

Theoretically, it can be used to derive linear filters for any linear
transform, as long as it is supplied with the input and outputs of a transform
pair. Previous works have relied on using theoretical transforms pairs commonly
found in tables of integral transforms. Here we follow a similar approach, but
we also propose a new method where filters could be derived from accurately
computed numerical transform pairs, such as those computed using quadrature
approaches. For example, the EM modeler \texttt{empymod} could be used to
compute transform pairs for a particular class of model and data parameters
(for example, a specific EM method and model of interest) using a slow but
highly accurate quadrature method. A good filter derived from the resulting
transform pairs could then be used for faster modeling of similar model and
data parameters.

The main features of \texttt{fdesign} and some differences between it and the
methods it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force), with an optional minimization routine for
    the best outcome of the brute-force step.
  \item \cite{GP.07.Kong} and \cite{GEO.12.Key} optimize a filter for the $J_0$
    filter, and then use the obtained shift and spacing value to calculate the
    $J_1$ values. The presented algorithm can optimize either $J_0$, $J_1$, or
    both in the optimization process. Five transform-pairs for each $J_0$ and
    $J_1$, and three transform pairs for each sine and cosine of often used
    functions are included in the routine, but any other transform pair can be
    provided as input.
  \item More complex or more specific numerical transforms  pairs  can be
    provided instead of theoretical transform pairs by using the modeler
    \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In  \texttt{fdesign}, you can define a
    relative error level that defines up to what error an obtained result is
    good or not. The optimization seeks to find smallest amplitude field that
    is accurately modeled (i.e., minimizing the amplitude), but it has
    additionally another mode, where it will maximize the abscissae $r$ of the
    right-hand-side of the transform pair instead of minimizing the amplitude.
    The two modes yield the same result in the general, simple case of fast
    decaying transform pairs. However, for some complex models and specifically
    for very high frequencies, maximizing $r$ will yield much better and more
    consistent results.
  \item The algorithm can solve under-, equal-, and over-determined linear
    systems.
  \item The real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

@@

Most of the articles mentioned in the review have detailed derivations of the
digital filter method. In this article we focus on the algorithm, and summarize
the theory only very briefly by following \cite{GEO.12.Key}. In
electromagnetics we often have to evaluate integrals of the form
%
\begin{equation}
  F(r) = \int^\infty_0 f(l)K(l r)\mr{d}l \ ,
  \label{eq:HankelInt}
\end{equation}
%
where $l$ and $r$ denote evaluation values, and $K$ is the kernel function. In
the specific case of the Hankel transform $l$ corresponds to wavenumber, $r$ to
offset, and $K$ to Bessel functions; in the case of the Fourier transform $l$
corresponds to frequency, $r$ to time, and $K$ to sine or cosine functions. In
both cases it is an infinite integral which numerical integration is very
time-consuming because of the slow decay of the kernel function and its
oscillatory behaviour.

@@

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. The source and receiver vertical separation is
1\,m. For the Fourier transform a 4096\,pt FFT was used with regularly spaced
frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up to 2048\,MHz. The
frequency result is multiplied with a Ricker wavelet with a center frequency of
250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is applied. Subplot (b)
shows the result when using the adaptive quadrature (QUAD) option of
\texttt{empymod}, which uses the QAGSE routine from the Fortran QUADPACK
library. Subplot (c) shows the result using DLF with the new 2001\,pt digital
filter. The calculation with DLF took under 9 minutes and is therefore roughly
80 times faster than the calculation with QUAD, which took roughly 11 hours and
27 minutes. Calculating the same model with QWE took 7 hours and 20 minutes.
However, QWE uses in roughly 1/3 of the calculation internally QUAD in this
example, see \cite{GEO.17.Werthmuller}. (Note that DLF was run in parallel
using 4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.)
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QUAD) or with (c) a 2001\,pt digital linear filter (b). The
  quadrature took about 11\,h 27\,min to calculate, whereas the filter took
  2\,min 10\,s in parallel on 4 nodes, hence under 9\,min in total.}
%

>>>>

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
%
% (1) PRINCIPAL OBJECTIVES AND SCOPE OF THE WORK
  The open-source code \texttt{fdesign} makes it possible to design digital
  linear filters for the Hankel and Fourier transforms used in potential-,
  diffusive-, and wavefield modeling. Digital filters can be derived for any
  electromagnetic method, such as methods in the diffusive limits (DC, CSEM) as
  well as methods using higher frequency content (GPR, acoustic and elastic
  wavefields).
%
% (2) METHODOLOGY
  The direct matrix inversion method is used for the derivation of the filter
  values, and a brute-force inversion is carried out over the defined spacing
  and shifting values of the filter basis. Included or user-provided
  theoretical transform pairs are used for the inversion. Alternatively, one
  can provide 1D subsurface models that will be computed with a precise
  quadrature method using the EM modeler \texttt{empymod} to generate
  numerical transform pairs.
%
% (3) RESULTS
  The comparison of the presented 201\,pt filter with previously presented
  filters shows that it performs better for some standard CSEM cases. The
  derivation of a longer 2001\,pt filter for a GPR example with 250\,MHz center
  frequency proves that the filter method works also for wave phenomena, not
  only for diffusive EM fields.
%
% (4) CONCLUSIONS
  The presented algorithm provides a tool to create problem specific digital
  filters. Such purpose-built filters can be made shorter and can speed up
  consecutive  potential-, diffusive-, and wavefield inversions.
%
\end{abstract}

@@

\cite{GP.71.Ghosh} states that the DLF idea is based on suggestions made four
decades earlier by \cite{PHY.33.Slichter} and \cite{GEO.40.Pekeris}, in that
``\emph{the kernel function is dependent only on the layer parameters, and an
expression relating it to the field measurements can be obtained by
mathematical processes.}'' However, until the introduction of DLF, these
suggestions found no widespread use, likely because of the missing computer
power to calculate the filter coefficients. He further states that credit goes
to \cite{BK.68.Koefoed, GP.70.Koefoed}, who retook the task of direct
interpretational methods with the introduction of raised kernel functions. DLF
is, as such, an improvement of that approach, providing a faster and simpler
method.

@@

The most well-known codes are likely Anderson's freely available ones.
\cite{USGS.73.Anderson} extends the method to transient responses, applying DLF
not only to the Hankel transform, but also to the Fourier transform. A
transient signal can therefore be obtained be applying twice a digital filter
to the wavenumber-frequency domain calculation. \cite{USGS.75.Anderson,
GEO.79.Anderson} presents improved filters for both Fourier and Hankel
transforms, introducing measures to significantly speed-up the calculation,
such as the lagged convolution or using the same abscissae for $J_0$ and $J_1$.
\cite{TMS.82.Anderson} and the included 801\,pt filter became the de facto
industry standard, to which subsequent filters were compared.
\cite{GEO.89.Anderson} presents a hybrid solution that uses either the DLF or a
much slower but highly accurate adaptive-quadrature approach, which can also be
used to measure the accuracy of the DLF. \cite{GEO.12.Key} presented codes for
comparing the DLF approach with a speed optimized quadrature method called
quadrature-with-extrapolation (QWE); despite its optimizations, QWE is still
not as fast as DLF, but it remains valuable as an independent tool for testing
the accuracy of particular filters for the DLF method. Other examples include
the codes by \cite{GP.75.Johansen}, an interactive system for interpretation of
resistivity soundings, and a tool to calculate filter coefficients by
\cite{GP.90.Christensen}. The latter is available upon request and was used,
for instance, in all the open-source modeling and inversion routines of CSIRO
in the Amira Project 223 \citep{ASEG.07.Raiche}.

@@

Theoretically, it can be used to derive linear filters for any linear
transform, as long as it is supplied with the inputs and outputs of a transform
pair. Previous works have relied on using theoretical transform pairs commonly
found in tables of integral transforms. Here we follow a similar approach, but
we also propose a new method where filters could be derived from accurately
computed numerical transform pairs, such as those computed using quadrature
approaches. For example, the EM modeler \texttt{empymod} could be used to
compute transform pairs for a particular class of model and data parameters
(for example, a specific EM method and model of interest) using a slow but
highly accurate quadrature method. A good filter derived from the resulting
transform pairs could then be used for faster modeling of similar model and
data parameters.

The main features of \texttt{fdesign} and some differences between it and the
methods it is based upon are:
\begin{itemize}
  \item The algorithm computes many different filters for various spacing and
    shift values (brute-force), with an optional minimization routine for
    the best outcome of the brute-force step.
  \item \cite{GP.07.Kong} and \cite{GEO.12.Key} optimize a filter for the $J_0$
    filter, and then use the obtained shift and spacing value to calculate the
    $J_1$ values. The presented algorithm can optimize either $J_0$, $J_1$, or
    both in the optimization process. Five transform-pairs for each $J_0$ and
    $J_1$, and three transform pairs for each sine and cosine of often used
    functions are included in the routine, but any other transform pair can be
    provided as input.
  \item More complex or more specific numerical transform  pairs  can be
    provided instead of theoretical transform pairs by using the modeler
    \texttt{empymod}.
  \item \cite{GP.07.Kong} defines a \emph{good} filter as one that recovers
    \emph{weak} diffusive EM fields. In  \texttt{fdesign}, you can define a
    relative error level that defines up to what error an obtained result is
    good or not. The optimization seeks to find the smallest amplitude field
    that is accurately modeled (i.e., minimizing the amplitude), but it has
    additionally another mode, where it will maximize the output abscissae $r$
    of the transform pair instead of minimizing the amplitude. The two modes
    yield the same result in the general, simple case of fast decaying
    transform pairs. However, for some complex models and specifically for very
    high frequencies, maximizing $r$ will yield much better and more consistent
    results.
  \item The algorithm can solve under-, equal-, and over-determined linear
    systems.
  \item The real or the imaginary part can be used for the inversion of complex
    transform pairs.
\end{itemize}

@@

Most of the articles mentioned in the review have detailed derivations of the
digital filter method. In this article we focus on the algorithm, and summarize
the theory only very briefly by following \cite{GEO.12.Key}. In
electromagnetics we often have to evaluate integrals of the form
%
\begin{equation}
  F(r) = \int^\infty_0 f(l)K(l r)\mr{d}l \ ,
  \label{eq:HankelInt}
\end{equation}
%
where $l$ and $r$ denote input and output evaluation values, respectively, and
$K$ is the kernel function. In the specific case of the Hankel transform $l$
corresponds to wavenumber, $r$ to offset, and $K$ to Bessel functions; in the
case of the Fourier transform $l$ corresponds to frequency, $r$ to time, and
$K$ to sine or cosine functions. In both cases it is an infinite integral which
numerical integration is very time-consuming because of the slow decay of the
kernel function and its oscillatory behaviour.

@@

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. The source and receiver vertical separation is
1\,m. For the Fourier transform a 4096\,pt FFT was used with regularly spaced
frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up to 2048\,MHz. The
frequency result is multiplied with a Ricker wavelet with a center frequency of
250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is applied. Subplot (b)
shows the result when using the adaptive quadrature (QUAD) option of
\texttt{empymod}, which uses the QAGSE routine from the Fortran QUADPACK
library. Subplot (c) shows the result using DLF with the new 2001\,pt digital
filter. The calculation with DLF took under 9 minutes and is therefore roughly
80 times faster than the calculation with QUAD, which took roughly 11 hours and
27 minutes. Calculating the same model with QWE took 7 hours and 20 minutes.
However, QWE uses in roughly 1/3 of the calculation internally QUAD in this
example, see \cite{GEO.17.Werthmuller}. (Note that DLF was run in parallel
using 4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.)
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QUAD) or with (c) a
  2001\,pt digital linear filter (b). QUAD took about 11\,h 27\,min to
  calculate, whereas DLF took 2\,min 10\,s in parallel on 4 nodes, hence under
  9\,min in total.}
%
